Parsing Postfix log files

### 2009-04-12 doc/logparser-acronyms.tex
Ran diction and made changes based on its suggestions.

\newglossaryentry{DNSBL}{
    type=\acronymtype,
    name={DNSBL},
    description={DNS Blacklist},
    first={DNS Blacklist (DNSBL)}
}
\newglossaryentry{DNSBL glossary}{
    name={DNS Blacklist},
    description={
        A DNS Blacklist is a simple collaborative anti-spam technique used
        to reject or penalise mail sent from mail servers believed to be
        the source of large volumes of spam.  The criteria used when
        deciding whether or not to include an IP address on a DNSBL vary
        widely between DNSBLs, so before using one it is very important to
        check their listing policies.  To use a DNSBL the mail server makes
        a DNS request incorporating the IP address of the client; if the
        requested hostname exists the client is on the DNSBL, and the mail
        server can decide what course of action to take.
    }
}

\newglossaryentry{API}{
    type=\acronymtype,
    name={API},
    description={Application Programming Interface},
    first={Application Programming Interface (API)}
}
\newglossaryentry{API glossary}{
    name={Application Programming Interface},
    description={
        One of the fundamental concepts when writing programs is the reuse
        of existing code, so that each new program does not reinvent
        existing wheels.  An API defines the interface provided to the user
        of the existing code, and acts as a contract between the user and
        the provider: if the user adheres to the API the provider
        guarantees it will work, but is free to change the implementation
        as long as the API is preserved.
    }
}

@@

\newglossaryentry{FA}{
    type=\acronymtype,
    name={FA},
    description={Finite Automata},
    first={Finite Automata (FA)}
}
\newglossaryentry{FA glossary}{
    name={Finite Automata},
    description={
        Finite automata are computing devices that accept or recognize
        regular languages.  They lack any form of storage, so while they
        can recognise languages such as ${(ab)}^{*}$ or
        $a^{*}{(b^{*}|c)}b$, they cannot count and so cannot recognise
        languages such as $a^{n}b^{n}$.
    }
}

\newglossaryentry{CFG}{
    type=\acronymtype,
    name={CFG},
    description={Context Free Grammar},
    first={Context Free Grammar (CFG)}
}
\newglossaryentry{CFG glossary}{
    name={Context Free Grammar},
    description={
        A CFG is a grammar in which every transition or production takes
        the form $L\rightarrow{}R$, where $L$ is a single non-terminal
        symbol, and $R$ is a (possibly empty) string of terminal and/or
        non-terminal symbols.  Every CFG can be recognised by a
        \acronym{PDA}, and \acronyms{PDA} cannot recognise more complicated
        grammars, so \acronyms{CFG} and \acronyms{PDA} are equivalent.
    }
}

>>>>

\newglossaryentry{DNSBL}{
    type=\acronymtype,
    name={DNSBL},
    description={DNS Blacklist},
    first={DNS Blacklist (DNSBL)}
}
\newglossaryentry{DNSBL glossary}{
    name={DNS Blacklist},
    description={
        A DNS Blacklist is a simple collaborative anti-spam technique used
        to reject or penalise mail sent from mail servers believed to be
        the source of large volumes of spam.  The criteria used when
        deciding if an IP address should be included on a DNSBL vary widely
        between DNSBLs, so before using one it is very important to check
        their listing policies.  To use a DNSBL the mail server makes a DNS
        request incorporating the IP address of the client; if the
        requested hostname exists the client is on the DNSBL, and the mail
        server can decide what course of action to take.
    }
}

\newglossaryentry{API}{
    type=\acronymtype,
    name={API},
    description={Application Programming Interface},
    first={Application Programming Interface (API)}
}
\newglossaryentry{API glossary}{
    name={Application Programming Interface},
    description={
        One of the fundamental concepts when writing programs is the reuse
        of existing code, so that each new program does not reinvent
        existing wheels.  An API defines the interface provided to the user
        of the existing code, and acts as a contract between the user and
        the provider: if the user adheres to the API the provider
        guarantees it will work, but is free to change the implementation
        if the API is preserved.
    }
}

@@

\newglossaryentry{FA}{
    type=\acronymtype,
    name={FA},
    description={Finite Automata},
    first={Finite Automata (FA)}
}
\newglossaryentry{FA glossary}{
    name={Finite Automata},
    description={
        Finite automata are computing devices that accept or recognize
        regular languages.  They lack any form of storage, so although they
        can recognise languages such as ${(ab)}^{*}$ or
        $a^{*}{(b^{*}|c)}b$, they cannot count and so cannot recognise
        languages such as $a^{n}b^{n}$.
    }
}

\newglossaryentry{CFG}{
    type=\acronymtype,
    name={CFG},
    description={Context Free Grammar},
    first={Context Free Grammar (CFG)}
}
\newglossaryentry{CFG glossary}{
    name={Context Free Grammar},
    description={
        A CFG is a grammar in which every transition or production takes
        the form $L\rightarrow{}R$, where $L$ is a single non-terminal
        symbol, and $R$ is a (possibly empty) string of terminal or
        non-terminal symbols.  Every CFG can be recognised by a
        \acronym{PDA}, and \acronyms{PDA} cannot recognise more complicated
        grammars, so \acronyms{CFG} and \acronyms{PDA} are equivalent.
    }
}

### 2009-04-12 doc/logparser-architecture.tex
Changes from proofreading of introduction chapter.
Consistent float placement.
Ran diction and made changes based on its suggestions.

\section{Parser Architecture and Design}

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that log files produced by Postfix may vary widely from host
to host, depending on the set of restrictions chosen by the administrator.
With this in mind, one of the architecture's design aims was to make adding
new rules to parse new inputs as easy as possible, to enable administrators
to properly parse their own log files.  The solution developed is to divide
the architecture into three parts: Framework, Actions, and Rules.  Each
will be discussed separately, but first an overview:

@@

There is some similarity between this architecture and William Wood's
\acronym{ATN}~\cite{atns,nlpip}, used in Computational Linguistics for
creating grammars to parse or generate sentences.  The resemblance between
the two (shown in \tableref{Similarities between ATN and this
architecture}) is accidental, but it is clear that the two different
approaches share a similar division of responsibilities, despite having
different semantics.

@@

\begin{table}[ht]
    \caption{Similarities between ATN and this architecture}
    \empty{}\label{Similarities between ATN and this architecture}
    \begin{tabular}[]{lll}
        \tabletopline{}%
        \acronym{ATN}   & Architecture  & Similarity                  \\
        \tablemiddleline{}%
        Networks        & Parser        & Determines the sequence
                                          of transitions              \\
                        &               & or actions that
                                          constitutes a valid input.  \\
        Transitions     & Actions       & Assemble data and
                                          impose conditions the       \\
                        &               & input must meet to be
                                          accepted as valid.          \\
        Abbreviations   & Rules         & Responsible for
                                          classifying input.          \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

    \item [Load and validate rules]  The framework loads the rules from
        wherever they are stored: a simple file, a database, or possibly
        even a web server or other network service --- though that would
        have serious security implications.  It validates each rule to
        hopefully catch problems as early in the parsing process as
        possible; the checks will be implementation-specific to some
        extent, but will generally include some of the following:

@@

    \item [Convert physical inputs to logical inputs] Each rule recognises
        a single input at a time: there is no facility for rules to consume
        more input or push unused input back onto the input stream,
        although actions may use cascaded parsing (explained in
        \sectionref{actions in architecture}) to effectively push input
        back onto the input stream.  A physical input (i.e.\ a single line
        read from the input stream) may contain multiple or partial logical
        inputs, and the framework must pre-process these to provide a
        logical input for the rules to recognise.  This simplifies the
        rules and actions considerably, at the cost of added complexity in
        the framework; during the design phase it was decided that it was
        easier to deal with the problem of parsing multiple or partial
        inputs once, rather than dealing with it in every rule and action.
        This is easy in the case of Postfix log files because there is a
        one-to-one mapping between physical and logical inputs; mapping
        between physical and logical inputs may be more difficult for other
        types of input.  Some input types may require pre-processing
        equivalent to parsing the physical inputs; in those cases the
        framework should behave similarly to other architectures: present
        the entire input stream at once, and discard the portion recognised
        by the successful rule.

@@

    \item [Specialised support functions]  Actions may need support or
        utility functions; the framework may be a good location for support
        functions, but if there is another way to make those functions
        available to all actions it may be preferable to use it instead,
        maintaining a clear separation of concerns.

@@

Each action is a separate procedure written to process a particular
category of input, e.g.\ rejecting a delivery attempt.  There may be many
input variants within one input category; in general each action will
handle one input category, with each rule recognising one input variant.
It is expected that parsers based on this architecture will have many
actions, and each action may be specified by many rules, with the aim of
having simple rules and tightly focused actions.  An action may need to
process different input variants in slightly different ways, but large
changes in processing indicate the need for a new action and a new category
of input; if an action becomes overly complicated it starts to turn into a
monolithic parser, with too much logic contained in a single procedure.

@@

Actions may return a modified input line to the framework that will be
parsed as if read from the input stream, allowing for a simplified version
of cascaded parsing~\cite{cascaded-parsing}.  This powerful facility allows
several rules and actions to parse a single input, potentially simplifying
both rules and actions.  A simple example is to have one rule and action
removing comments from inputs, so that other rules and actions do not
have to handle comments at all; obviously if comment characters can be
escaped or embedded in quoted strings the implementation must be careful
not to remove those.  For some inputs this type of pre-processing can
greatly simplify parsing, echoing the simplification provided by the
framework presenting rules and actions with logical inputs rather than
physical inputs.  A more complex example use of cascaded parsing is
evaluating simple arithmetic expressions, where sub-expressions enclosed in
parentheses must be evaluated first; cascaded parsing can be used to parse
and evaluate the sub-expressions, substituting the result into the original
expression for subsequent re-evaluation.  Actions do not need to be
specially registered with the framework or be declared in a particular way
to use cascaded parsing: actions that do not use cascaded parsing will
return nothing, those that do will simply return a string to be parsed.

@@

\begin{table}[ht]
    \caption{Rules to parse C-style comments}
    \empty{}\label{Rules to parse C-style comments}
    \begin{tabular}{llll}
        \tabletopline{}%
        No.   & Regex             & Condition             & Variable Changes    \\
        \tablemiddleline{}%
        1     & \verb!/\*.*?\*/!  & expecting == code     & expecting = code      \\
        2     & \verb!/\*.*!      & expecting == code     & expecting = comment   \\
        3     & \verb!.*\*/!      & expecting == comment  & expecting = code      \\
        4     & \verb!.*!         & expecting == comment  & expecting = comment   \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

Overlapping rules can be detected by visual inspection, or a program may be
written to analyse the rule's regexes.  Traditional regexes are equivalent
in computational power to \acronym{FA} and can be converted to
\acronym{FA}, so regex overlap can be detected by finding a non-empty
intersection of two \acronym{FA}\@.  Perl 5.10 regexes~\cite{perlre} are
more powerful than traditional regexes: it is possible to match correctly
balanced brackets nested to an arbitrary depth, e.g.\
\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!  matches
\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
regex engine to maintain state on a stack, so Perl 5.10 regexes are
equivalent in computational power to \acronym{PDA}; detecting overlap may
require calculating the intersection of two \acronym{PDA} instead of two
\acronym{FA}.  \acronym{PDA} intersection is not closed, i.e.\ the result
cannot always be implemented using a \acronym{PDA}, so intersection may be
intractable in some cases, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating
$\frac{n\left(n-1\right)}{2}$ intersections, resulting in
$O\left(n^{2}x\right)$ complexity, where $x$ is the cost of calculating
\acronym{FA} or \acronym{PDA} intersection.  This is certainly not a task
to be performed every time a parser runs: naive detection of overlap
amongst \parsernames{} \numberOFrules{} rules would require calculating
\numberOFruleINTERSECTIONS{} intersections.

When detecting overlap any conditions attached to rules must be taken into
account, because two rules whose regexes overlap may have conditions
attached which prevent the rules overlapping.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlap between the regexes of each pair of rules
with overlapping conditions.  Depending on the conditions employed a rule
may overlap with multiple rules, and rule overlap is not transitive, e.g.\
given these three artificial conditions:

@@

Once conditions pass a certain level of complexity determining whether two
conditions overlap becomes intractable, because it requires so much
knowledge of other rules and possibly even actions.  E.g.\ given two rules
with conditions \verb!verbose == true! and \verb!silent == true!, logically
these rules should not overlap, yet there is nothing to stop both variables
being set to true by one or more rules or actions.  If variables used in
conditions can be set by actions, determining whether or not two conditions
overlap is impossible: the Halting Problem show that it is impossible for
one program to answer questions about another program's behaviour if the
inquiring program is implemented on a computational machine of the equal or
lesser power to the program being inquired into.

@@

It is possible to define pathological regexes which fall into two main
categories: regexes that match inputs they should not, and regexes that
consume excessive amounts of CPU time during matching.  Defining a regex
that matches inputs it should not is trivial: \verb!/^/! matches the start
of every input.  This regex would be found by a tool that detects
overlapping rules, and would easily be noticed by visual inspection, but
more complex regexes would be harder to spot.  Regexes that match more than
they should are a problem not because of excessive resource usage, but
because they may prevent the correct rule from recognising the input.  If
an adaptive ordering system is used that prioritises rules that frequently
recognise inputs (see \sectionref{rule ordering for efficiency}), then a
rule with a regex that matches inputs it should not may be promoted up
through the list, displacing an increasing number of correct rules as it
moves.

Usually excessive CPU time is consumed when a regex with a lot of
backtracking, caused by alteration and/or nested quantifiers, fails to
match; successful matching is generally quite fast with such regexes, so
problematic regexes may go unnoticed for some time.  E.g.\ with most regex
engines matching double quoted strings with \verb!"([^"\\]+|\\.)*"! is very
fast when there is a match, but when the match fails its computational
complexity for a string of length $n$ is $O(2^{n})$;
see~\cite{mastering-regular-expressions} for in-depth discussion of nested
quantifiers, backtracking, alteration, and capturing groups.  Pathological
regexes which consume excessive CPU time can be difficult to detect, both
by visual inspection and by machine inspection, but if a regex is converted
to a \acronym{FA} or the internal representation used by the regex engine,
it may be possible to determine if there are nested quantifiers or other
troublesome constructs present.  Modern regex engines have addressed many
of these problems, e.g.\ the regex to match double quoted strings given
above fails immediately with Perl 5.10, regardless of the input length,
because the regex engine looks for both of the required double quotes
first.  Similarly Perl 5.10's regex engine optimises alterations starting
with literal text into a trie, which has matching time proportional to the
length of the alternatives, rather than the number of alternatives.  Perl
regexes can use \verb!(?>pattern)!, which matches \verb!pattern! the first
time the regex engine passes over it, and does not change what it
originally matched if the regex engine backtracks over it, so troublesome
regexes can use this to reduce the impact of backtracking; Prolog users
will notice a similarity to the \verb'!' (cut) operator.  A presentation
showing some of Perl 5.10's new regex features is available at
\urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.

Conditions can vary in complexity from simple equality through to a
Turing-equivalent language, so enumerating pathological conditions is
difficult if not pointless.  Conditions that examine the input using
regexes can suffer from the same problems as rule regexes; conditions that
check variables in uncomplicated ways may exhibit unexpected or incorrect
behaviour, but are unlikely to exhibit pathological behaviour.  Deciding
whether a more complex condition's behaviour is pathological or simply a
bug is difficult and to some extent is a matter of opinion.  When this
architecture has received more widespread usage consensus should be reached
on the topic of pathological conditions.

@@

This chapter has presented the parser architecture developed for this
project.  It started with a high level view of the architecture, describing
how it achieves its design aim of being easily extensible for users, and
the advantages that being easily extensible brings to parser authors too.
The three main components of the architecture were documented in detail,
explaining each component's responsibilities and the functionality it
provides, plus any difficulties associated with the components.  The
framework provides a number of support functions and manages the parsing
process, enabling simple rules and actions to be written.  The actions are
quite easy to understand, because the architecture does not impose any
structure or requirements upon them: parser authors are free to do anything
they want within an action.  The architecture's support for cascaded
parsing was described in the actions section, with an example to illustrate
how it can be useful for general parsing.  The rules section was the
longest, because although the rules appear conceptually simple ---
recognise an input and specify the action that handles that input --- they
have subtle behaviour that needs to be clearly explained.  When extending a
ruleset a decision needs to be taken about whether the input should be
recognised by extending an existing rule, by adding a new rule to an
existing rule category, or by adding a new rule category, rule, and action.
Rules can have conditions attached to them, restricting the set of rules
used to recognise an input; the complexity of the conditions used greatly
influences the difficulty of writing a correct ruleset or understanding and
extending an existing ruleset.  Overlapping rules are frequently a
requirement in a parser, and their use can greatly simplify some rules, but
they can be a source of bugs because they can recognise inputs
unexpectedly.  The framework does not try to detect overlapping rules,
because overlap amongst rules may be valid and is quite often intentional;
that responsibility falls to the author of the rules.  The difficulty
involved in detecting overlap is proportional to the complexity of a
ruleset's regexes and conditions, and may be possible for a human yet
intractable or impossible for a program.  The rules section finishes with a
discussion of pathological rules, concentrating on pathological regexes.

>>>>

\section{Architecture Overview}

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that log files produced by Postfix may vary widely from host
to host, depending on the set of restrictions chosen by the administrator.
With this in mind, one of the architecture's design aims was to make adding
new rules to parse new inputs as effortless as possible, to enable
administrators to properly parse their own log files.  The solution
developed is to divide the architecture into three parts: Framework,
Actions, and Rules.  Each will be discussed separately, but first an
overview:

@@

Some similarity exists between this architecture and William Wood's
\acronym{ATN}~\cite{atns,nlpip}, used in Computational Linguistics for
creating grammars to parse or generate sentences.  The resemblance between
the two (shown in \tableref{Similarities between ATN and this
architecture}) is accidental, but clearly the two different approaches
share a similar division of responsibilities, despite having different
semantics.

@@

\begin{table}[thbp]
    \caption{Similarities between ATN and this architecture}
    \empty{}\label{Similarities between ATN and this architecture}
    \begin{tabular}[]{lll}
        \tabletopline{}%
        \acronym{ATN}   & Architecture  & Similarity                  \\
        \tablemiddleline{}%
        Networks        & Parser        & Determines the sequence
                                          of transitions              \\
                        &               & or actions that
                                          constitutes a valid input.  \\
        Transitions     & Actions       & Assemble data and
                                          impose conditions the       \\
                        &               & input must meet to be
                                          accepted as valid.          \\
        Abbreviations   & Rules         & Responsible for
                                          classifying input.          \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

    \item [Load and validate rules]  The framework loads the rules from
        wherever they are stored: a simple file, a database, or possibly
        even a web server or other network service --- though that would
        have serious security implications.  It validates each rule to
        catch problems as early in the parsing process as possible; the
        checks will be implementation-specific to some extent, but will
        generally include some of the following:

@@

    \item [Convert physical inputs to logical inputs] Each rule recognises
        a single input at a time: there is no facility for rules to consume
        more input or push unused input back onto the input stream,
        although actions may use cascaded parsing (explained in
        \sectionref{actions in architecture}) to effectively push input
        back onto the input stream.  A physical input (i.e.\ a single line
        read from the input stream) may contain multiple or partial logical
        inputs, and the framework must pre-process these to provide a
        logical input for the rules to recognise.  This simplifies the
        rules and actions considerably, at the cost of added complexity in
        the framework; during the design phase it was decided that it was
        easier to deal with the problem of parsing multiple or partial
        inputs once, rather than dealing with it in every rule and action.
        This is trivial for Postfix log files because they have a
        one-to-one mapping between physical and logical inputs; mapping
        between physical and logical inputs may be more difficult for other
        types of input.  Some input types may require pre-processing
        equivalent to parsing the physical inputs; in those cases the
        framework should behave similarly to other architectures: present
        the entire input stream at once, and discard the portion recognised
        by the successful rule.

@@

    \item [Specialised support functions]  Actions may need support or
        utility functions; the framework may be a good location for support
        functions, but if another way exists to make those functions
        available to all actions it may be preferable to use it instead,
        maintaining a clear separation of concerns.

@@

Each action is a separate procedure written to process a particular
category of input, e.g.\ rejecting a delivery attempt.  There may be many
input variants within one input category; in general each action will
handle one input category, with each rule recognising one input variant.
It is anticipated that parsers based on this architecture will have many
actions, and each action may be specified by many rules, with the aim of
having simple rules and tightly focused actions.  An action may need to
process different input variants in slightly different ways, but large
changes in processing indicate the need for a new action and a new category
of input; if an action becomes overly complicated it starts to turn into a
monolithic parser, with too much logic contained in a single procedure.

@@

Actions may return a modified input line to the framework that will be
parsed as if read from the input stream, allowing for a simplified version
of cascaded parsing~\cite{cascaded-parsing}.  This powerful facility allows
several rules and actions to parse a single input, potentially simplifying
both rules and actions.  A simple example is to have one rule and action
removing comments from inputs, so that other rules and actions do not
have to handle comments at all; obviously if comment characters can be
escaped or embedded in quoted strings the implementation must be careful
not to remove those.  For some inputs this kind of pre-processing can
greatly simplify parsing, echoing the simplification provided by the
framework presenting rules and actions with logical inputs rather than
physical inputs.  A more complex example use of cascaded parsing is
evaluating simple arithmetic expressions, where sub-expressions enclosed in
parentheses must be evaluated first; cascaded parsing can be used to parse
and evaluate the sub-expressions, substituting the result into the original
expression for subsequent re-evaluation.  Actions do not need to be
specially registered with the framework or be declared in a particular way
to use cascaded parsing: actions that do not use cascaded parsing will
return nothing, those that do will simply return a string to be parsed.

@@

\begin{table}[thbp]
    \caption{Rules to parse C-style comments}
    \empty{}\label{Rules to parse C-style comments}
    \begin{tabular}{llll}
        \tabletopline{}%
        No.   & Regex             & Condition             & Variable Changes    \\
        \tablemiddleline{}%
        1     & \verb!/\*.*?\*/!  & expecting == code     & expecting = code      \\
        2     & \verb!/\*.*!      & expecting == code     & expecting = comment   \\
        3     & \verb!.*\*/!      & expecting == comment  & expecting = code      \\
        4     & \verb!.*!         & expecting == comment  & expecting = comment   \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

Overlapping rules can be detected by visual inspection, or a program may be
written to analyse the rule's regexes.  Traditional regexes are equivalent
in computational power to \acronym{FA} and can be converted to
\acronym{FA}, so regex overlap can be detected by finding a non-empty
intersection of two \acronym{FA}\@.  Perl 5.10 regexes~\cite{perlre} are
more powerful than traditional regexes: it is possible to match correctly
balanced brackets nested to an arbitrary depth, e.g.\
\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!  matches
\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
regex engine to maintain state on a stack, so Perl 5.10 regexes are
equivalent in computational power to \acronym{PDA}; detecting overlap may
require calculating the intersection of two \acronym{PDA} instead of two
\acronym{FA}.  \acronym{PDA} intersection is not closed, i.e.\ the result
cannot always be implemented using a \acronym{PDA}, so intersection may be
intractable for some combinations, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating
$\frac{n\left(n-1\right)}{2}$ intersections, resulting in
$O\left(n^{2}x\right)$ complexity, where $x$ is the cost of calculating
\acronym{FA} or \acronym{PDA} intersection.  This is certainly not a task
to be performed every time a parser runs: naive detection of overlap
amongst \parsernames{} \numberOFrules{} rules would require calculating
\numberOFruleINTERSECTIONS{} intersections.

When detecting overlap any conditions attached to rules must be taken into
account, because two rules whose regexes overlap may have conditions
attached that prevent the rules overlapping.  A less naive approach to
overlap detection would first check for overlapping conditions amongst
rules, and then check for overlap between the regexes of each pair of rules
with overlapping conditions.  Depending on the conditions employed a rule
may overlap with multiple rules, and rule overlap is not transitive, e.g.\
given these three artificial conditions:

@@

Once conditions pass a certain level of complexity determining if two
conditions overlap becomes intractable, because it requires so much
knowledge of other rules and possibly even actions.  E.g.\ given two rules
with conditions \verb!verbose == true! and \verb!silent == true!, logically
these rules should not overlap, yet there is nothing to stop both variables
being set to true by one or more rules or actions.  If variables used in
conditions can be set by actions, determining if two conditions overlap is
impossible: the Halting Problem show that it is impossible for one program
to answer questions about another program's behaviour if the inquiring
program is implemented on a computational machine of the equal or lesser
power to the program being inquired into.

@@

It is possible to define pathological regexes, which fall into two main
categories: regexes that match inputs they should not, and regexes that
consume excessive amounts of CPU time during matching.  Defining a regex
that matches inputs it should not is trivial: \verb!/^/! matches the start
of every input.  This regex would be found by a tool that detects
overlapping rules, and would easily be noticed by visual inspection, but
more complex regexes would be harder to spot.  Regexes that match more than
they should are a problem not because of excessive resource usage, but
because they may prevent the correct rule from recognising the input.  If
an adaptive ordering system is used that prioritises rules that frequently
recognise inputs (see \sectionref{rule ordering for efficiency}), then a
rule with a regex that matches inputs it should not may be promoted up
through the list, displacing an increasing number of correct rules as it
moves.

Usually excessive CPU time is consumed when a regex fails to match, and the
regex engine backtracks many times because of alteration or nested
quantifiers; successful matching is generally quite fast with such regexes,
so problematic regexes may go unnoticed for some time.  E.g.\ with most
regex engines matching double quoted strings with \verb!"([^"\\]+|\\.)*"!
is very fast when there is a match, but when the match fails its
computational complexity for a string of length $n$ is $O(2^{n})$;
see~\cite{mastering-regular-expressions} for in-depth discussion of nested
quantifiers, backtracking, alteration, and capturing groups.  Pathological
regexes that consume excessive CPU time can be difficult to detect, both by
visual inspection and by machine inspection, but if a regex is converted to
a \acronym{FA} or the internal representation used by the regex engine, it
may be possible to determine if nested quantifiers or other troublesome
constructs are present.  Modern regex engines have addressed many of these
problems, e.g.\ the regex to match double quoted strings given above fails
immediately with Perl 5.10, regardless of the input length, because the
regex engine looks for both of the required double quotes first.  Similarly
Perl 5.10's regex engine optimises alterations starting with literal text
into a trie, which has matching time proportional to the length of the
alternatives, rather than the number of alternatives.  Perl regexes can use
\verb!(?>pattern)!, which matches \verb!pattern! the first time the regex
engine passes over it, and does not change what it originally matched if
the regex engine backtracks over it, so troublesome regexes can use this to
reduce the impact of backtracking; Prolog users will notice a similarity to
the \verb'!' (cut) operator.  A presentation showing some of Perl 5.10's
new regex features is available at
\urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.

Conditions can vary in complexity from simple equality through to a
Turing-equivalent language, so enumerating pathological conditions is
difficult if not pointless.  Conditions that examine the input using
regexes can suffer from the same problems as rule regexes; conditions that
check variables in uncomplicated ways may exhibit unexpected or incorrect
behaviour, but are unlikely to exhibit pathological behaviour.  Deciding if
a more complex condition's behaviour is pathological or simply a bug is
difficult and to some extent is a matter of opinion.  When this
architecture has received more widespread usage consensus should be reached
on the topic of pathological conditions.

@@

This chapter has presented the parser architecture developed for this
project.  It started with a high level view of the architecture, describing
how it achieves its design aim of being easily extensible for users, and
the advantages that being easily extensible brings to parser authors too.
The three main components of the architecture were documented in detail,
explaining each component's responsibilities and the functionality it
provides, plus any difficulties associated with the components.  The
framework provides several support functions and manages the parsing
process, enabling simple rules and actions to be written.  The actions are
simple to understand, because the architecture does not impose any
structure or requirements upon them: parser authors are free to do anything
they want within an action.  The architecture's support for cascaded
parsing was described in the actions section, with an example to illustrate
how it can be useful for general parsing.  The rules section was the
longest, because although the rules appear conceptually simple ---
recognise an input and specify the action that handles that input --- they
have subtle behaviour that needs to be clearly explained.  When extending a
ruleset a decision needs to be taken about whether the input should be
recognised by extending an existing rule, by adding a new rule to an
existing rule category, or by adding a new rule category, rule, and action.
Rules can have conditions attached to them, restricting the set of rules
used to recognise an input; the complexity of the conditions used greatly
influences the difficulty of writing a correct ruleset or understanding and
extending an existing ruleset.  Overlapping rules are frequently a
requirement in a parser, and their use can greatly simplify some rules, but
they can be a source of bugs because they can recognise inputs
unexpectedly.  The framework does not try to detect overlapping rules,
because overlap amongst rules may be valid and is quite often intentional;
that responsibility falls to the author of the rules.  The difficulty
involved in detecting overlap is proportional to the complexity of a
ruleset's regexes and conditions, and may be possible for a human yet
intractable or impossible for a program.  The rules section finishes with a
discussion of pathological rules, concentrating on pathological regexes.
